---
title: "Variable Selection (linear models)"
output:
  html_document:
    df_print: paged
    keep_md: true
---

In this [R Notebook](http://rmarkdown.rstudio.com/r_notebooks.html) we will test exercise and eval some model selection strategies involving linear regration.

## Data set

We will use Boston dataset in the [MASS package](https://cran.r-project.org/web/packages/MASS/MASS.pdf) with the housing values in suburbs of Boston, the dataframe has 506 rows and 14 columns.

```{r setup, warning=FALSE, error=FALSE, message=FALSE}
# basic library
library(MASS)
library(tidyverse)
library(ggplot2)
library(grid)
library(gridExtra)

# dataset
bd <- Boston
str(bd)
```

We'll try to stablish a model for variable `medv` (median value of owner-occupied homes in \$1000), using a linear/polynomial model of other variables. In this case, the number of variables is low (13) so it's possible see an overview of the relationship between them and the `medv`.

```{r featurePlot}

# what are the predictors?
features <- names(bd)
features <- features[features!="medv"] # removing target (y)

map(features,function(feat,data=bd){
  ggplot(data, aes_string(x=feat, y="medv")) +
    geom_point() + theme_bw() + stat_smooth(method="lm") +  ylab("")
}) -> featPlots

marrangeGrob(featPlots, nrow=3, ncol=5)

```

